### 2014-07-20 docs/paper/text/solution.tex
minor fix to the overlapping case
minor fix to the non-overlapping case ILP formulation text

\paragraph*{Problem.} \emph{Find a true partitioning of attributes that
minimizes the query I/O and bounds the storage cost by some upper limit.}

@@

Each of these variables serve an important purpose:
\begin{itemize}
\item $x$s define the attribute to partition assignment.
\item $y$s help formulate the objective function's I/O contributions
that are on a per-partition basis.
\item $z$s help formulate the objective function's I/O contributions
that are on a per-attribute basis.
\item $u$s help formulate the storage overhead requirement.
\end{itemize}

@@

We are now ready to state the ILP formulation. We start with the objective
function, that is the total query I/O cost:
\begin{eqnarray}
\sum_{q\in Q} \sum_{p=1}^{k} && (16\cdot c_e(B) + 12\cdot c_n(B))\cdot
y_{p,q}\nonumber\\ 
&+& \sum_{a\in A} s(a)\cdot c_e(B)\cdot z_{a,p,q}\label{eq:no-obj}
\end{eqnarray}

In Eq.~\ref{eq:no-obj}, we simply sum for each query and each partiton to add
the I/O cost of reading in the structural information found in a sub-block, if
the partition is used by the query. We then sum over each attribute as well,
and add the I/O cost of reading in the attributes. Note that $z_{a,p,q}$ could
have been  replaced with $x_{a,p}\cdot y_{p,q}$, but that would make the
objective function non-linear.

Our first constraint is that each attribute gets assigned to one and only one
partition. Formally:
\begin{eqnarray}
\forall_{a\in A}, \sum_{p=1}^{k} x_{a,p} = 1
\end{eqnarray}

Our second constraint is that, if a query contains an attribute assigned to a
partition, then that partition is used by the query. In essence, we want to
state: $\forall_{\{p,q\}\in [1..k]\times Q}, y_{p,q} = \mathbf{1}(\sum_{a\in
A} q(a)\cdot x_{a,p}>0)$. In order to formulate this we use the following ILP 
construction:

Assume we have two variables, $\beta_1$ and $\beta_2$, where $\beta_2\in[0,1]$
and $\beta_1\geq 0$. We want to implement the following constraint: $\beta_2 =
\mathbf{1}(\beta_1 > 0)$.

This could be expressed as a linear constraint as follows, where $K$ is a
large constant guaranteed to be larger than $\beta_1$ for practical purposes:
\begin{eqnarray}
&& \beta_1 - \beta_2 \geq 0\nonumber\\
&& K\cdot\beta_2 - \beta_1 \geq 0\nonumber
\end{eqnarray}

@@

Our third constraint is that, if an attribute $a$ is assigned to a partition
$p$, and partition $p$ is used by a query $q$, then the corresponding $z$
variable must be set to $1$. It must be $0$ in all other cases. That is, we
want: $\forall_{\{a,p,q\}\in A\times [1..k]\times Q},
z_{a,p,q}=\mathbf{1}(x_{a,p} = y_{p,q} = 1)$. We express this as a linear 
constraint, as follows:
\begin{eqnarray}
\forall_{\{a,p,q\}\in A\times [1..k]\times Q},
    && z_{a,p,q} - (x_{a,p} + y_{p,q}) \geq -1\label{eq:no-z}
\end{eqnarray}

In Eq.~\ref{eq:no-z}, when the $x$ and $y$ variable are both $1$, the  $z$
variable is simply forced to be $1$. Otherwise, the $z$ variable can be either
$0$ or $1$, but since the $z$ variables appear in the objective fucntion as
positive terms, the solver will set them to $0$, which is what we want.

Our fourth constraint is that, if a partition is non-empty, then its
corresponding $u$ variable must be set to $0$. It must be $0$ otherwise. In
other words,  we want $u_p = \mathbf{1}(\sum_{a\in A} x_{a,p}>0)$. This is
expressed as linear constraints, as follows:
\begin{eqnarray}
\forall_{p\in[1..k]},
    && \sum_{a\in A} x_{a,p} - u_p \geq 0 \nonumber\\
\forall_{p\in[1..k]},
    && K\cdot u_p - \sum_{a\in A} x_{a,p} \geq 0 \label{eq:no-u}
\end{eqnarray}

@@

\subsubsection{Integer Linear Program Formulation}
% NOTE: In the simulation we will have query weights that will factor into the objective function
\begin{eqnarray}
\text{minimize}  
    \sum_{q\in Q} \sum_{p=1}^{k} && (16\cdot c_e(B) + 12\cdot c_n(B))\cdot y_{p,q}\nonumber\\
    &+& \sum_{a\in A} s(a)\cdot c_e(B)\cdot z_{a,p,q} \nonumber\\
\text{subject to}&&\nonumber\\
\forall_{a\in A}, 
    && \sum_{p=1}^{k} x_{a,p} > 0\nonumber\\
\forall_{a\in A}, 
    && \sum_{p=1}^{k} x_{a,p} \leq k\nonumber\\
\forall_{\{a,q\}\in A\times Q},
    &&  \sum_{p=1}^{k} z_{a,p,q} \geq q(a) \nonumber\\
\forall_{\{p,q\}\in [1..k]\times Q}, 
    &&  \sum_{a\in A} z_{a,p,q} - y_{p,q} \geq 0 \nonumber\\
\forall_{\{p,q\}\in [1..k]\times Q}, 
    &&  K\cdot y_{p,q} - \sum_{a\in A} z_{a,p,q}  \geq 0 \nonumber\\
\forall_{\{a,p,q\}\in A\times [1..k]\times Q}, 
    && z_{a,p,q} - (y_{p,q} + q(a)) \geq -1 \nonumber\\
\forall_{p\in[1..k]},
    && \sum_{a\in A} x_{a,p} - u_p \geq 0 \nonumber\\
\forall_{p\in[1..k]},
    && K\cdot u_p - \sum_{a\in A} x_{a,p} \geq 0 \nonumber\\    
&& \sum_{p=1}^{k} u_p -1 \leq \frac{\alpha}
  {1-\frac{c_e(B)\cdot \sum_{a\in A} s(a)}{s(B)}} \nonumber
\end{eqnarray}

>>>>

\paragraph*{Problem.$\,$} \emph{Find a true partitioning of attributes that
minimizes the query I/O and bounds the storage cost by some upper limit.}

@@

Each of these variables serve a purpose:
\begin{itemize}
\item $x$s define the attribute-to-partition assignment.
\item $y$s help formulate the query I/O contribution
of partitions, excluding their assigned attributes.
\item $z$s help formulate the query I/O contribution
of partitions, only considering their assigned attributes.
\item $u$s help formulate the storage overhead requirement.
\end{itemize}

@@

We are now ready to state the ILP formulation. We start with the objective
function, that is the total query I/O:
\begin{eqnarray}
\sum_{q\in Q} \sum_{p=1}^{k} && (16\cdot c_e(B) + 12\cdot c_n(B))\cdot
y_{p,q}\nonumber\\ 
&+& \sum_{a\in A} s(a)\cdot c_e(B)\cdot z_{a,p,q}\label{eq:no-obj}
\end{eqnarray}

In Eq.~\ref{eq:no-obj}, we simply sum for each query and each partition, and
add the I/O cost of reading in the structural information found in a
sub-block, if the partition is used by the query. We then sum over each
attribute as well, and add the I/O cost of reading in the attributes. Note
that $z_{a,p,q}$ could have been  replaced with $x_{a,p}\cdot y_{p,q}$, but
that would make the objective function non-linear.

We are now ready to state our constraints. Our first constraint is that, each
attribute must be assigned to a single partition. Formally:
\begin{eqnarray}
\forall_{a\in A}, \sum_{p=1}^{k} x_{a,p} = 1
\end{eqnarray}

Our second constraint is that, if a query $q$ contains an attribute $a$
assigned to a partition $p$, then partition $p$ is used by the query, i.e.,
$y_{p,q}=1$. In essence, we want to state: $\forall_{\{p,q\}\in [1..k]\times
Q}, y_{p,q} = \mathbf{1}(\sum_{a\in A} q(a)\cdot x_{a,p}>0)$.

In order to formulate this constraint, we use the following ILP  construction:
Assume we have two variables, $\beta_1$ and $\beta_2$, where $\beta_2\in[0,1]$
and $\beta_1\geq 0$. We want to implement the following constraint: $\beta_2 =
\mathbf{1}(\beta_1 > 0)$. This could be expressed as a linear constraint as
follows, where $K$ is a large constant guaranteed to be larger than $\beta_1$
for practical purposes:
\begin{eqnarray}
&& \beta_1 - \beta_2 \geq 0\nonumber\\
&& K\cdot\beta_2 - \beta_1 \geq 0\nonumber
\end{eqnarray}

@@

Our third constraint is that, if an attribute $a$ is assigned to a partition
$p$, and partition $p$ is used by a query $q$, then the corresponding $z$
variable must be set to $1$. That is, we want: $\forall_{\{a,p,q\}\in A\times
[1..k]\times Q}, z_{a,p,q}=\mathbf{1}(x_{a,p} = y_{p,q} = 1)$. We express this
as a linear  constraint, as follows:
\begin{eqnarray}
\forall_{\{a,p,q\}\in A\times [1..k]\times Q},
    && z_{a,p,q} - (x_{a,p} + y_{p,q}) \geq -1\label{eq:no-z}
\end{eqnarray}

In Eq.~\ref{eq:no-z}, when the $x$ and $y$ variables are both $1$, the  $z$
variable is simply forced to be $1$. Otherwise, the $z$ variable can be either
$0$ or $1$, but since the $z$ variables appear in the objective fucntion as
positive terms, the solver will set them to $0$, which is what we want.

Our fourth constraint is that, if a partition is non-empty, then its
corresponding $u$ variable must be set to $0$. In other words,  we want
$\forall_{p\in[1..k]}, u_p = \mathbf{1}(\sum_{a\in A} x_{a,p}>0)$. This is
expressed as linear constraints, as follows:
\begin{eqnarray}
\forall_{p\in[1..k]},
    && \sum_{a\in A} x_{a,p} - u_p \geq 0 \nonumber\\
\forall_{p\in[1..k]},
    && K\cdot u_p - \sum_{a\in A} x_{a,p} \geq 0 \label{eq:no-u}
\end{eqnarray}

@@

\subsubsection{Integer Linear Program Formulation}
% NOTE: In the simulation we will have query weights that will factor into the objective function
\begin{eqnarray}
\text{minimize}  
    \sum_{q\in Q} \sum_{p=1}^{k} && (16\cdot c_e(B) + 12\cdot c_n(B))\cdot y_{p,q}\nonumber\\
    &+& \sum_{a\in A} s(a)\cdot c_e(B)\cdot z_{a,p,q} \nonumber\\
\text{subject to}&&\nonumber\\
\forall_{a\in A}, 
    && \sum_{p=1}^{k} x_{a,p} > 0\nonumber\\
\forall_{a\in A}, 
    && \sum_{p=1}^{k} x_{a,p} \leq k\nonumber\\
\forall_{\{a,q\}\in A\times Q},
    &&  \sum_{p=1}^{k} z_{a,p,q} \geq q(a) \nonumber\\
\forall_{\{p,q\}\in [1..k]\times Q}, 
    &&  \sum_{a\in A} z_{a,p,q} - y_{p,q} \geq 0 \nonumber\\
\forall_{\{p,q\}\in [1..k]\times Q}, 
    &&  K\cdot y_{p,q} - \sum_{a\in A} z_{a,p,q}  \geq 0 \nonumber\\
\forall_{\{a,p,q\}\in A\times [1..k]\times Q}, 
    && z_{a,p,q} - (x_{a,p} + y_{p,q} + q(a)) \geq -2 \nonumber\\
\forall_{p\in[1..k]},
    && \sum_{a\in A} x_{a,p} - u_p \geq 0 \nonumber\\
\forall_{p\in[1..k]},
    && K\cdot u_p - \sum_{a\in A} x_{a,p} \geq 0 \nonumber\\    
&& \sum_{p=1}^{k} u_p -1 \leq \frac{\alpha}
  {1-\frac{c_e(B)\cdot \sum_{a\in A} s(a)}{s(B)}} \nonumber
\end{eqnarray}
